---
title: "Modeling spatially varying gene correlation across 1D curve"
author: >
  Chenxin (Flora) Jiang<br>
  Department of Statistics and Data Science, University of California, Los Angeles<br>
  <a href="mailto:cflorajiang@g.ucla.edu">cflorajiang@g.ucla.edu</a>
date: "`r format(Sys.Date(), '%d %B %Y')`"
output: 
  BiocStyle::html_document:
    number_sections: false
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
vignette: >
  %\VignetteIndexEntry{spCorr-1D}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
tools::R_user_dir("spCorr", which = "cache")
```

```{r setup, message=FALSE, warning=FALSE, results='hide'}
library(spCorr)
library(Seurat)
library(purrr)
library(tibble)
library(ggplot2)
library(dplyr)
library(tidyr)
library(viridis)
library(MorphoGAM)
library(stats)
library(utils)
```

## Introduction

This tutorial demonstrates how to use **spCorr** to
(i) infer spot-level gene–gene correlations for specified gene pairs, and
(ii) identify gene pairs exhibiting *spatially varying correlation (SVC)* patterns across a 1D spatial curve.

## Prepare input data

In this example, we analyze a subset of the **mouse brain cortex** dataset generated using the **10x Xenium** platform. The dataset is provided by 10x Genomics as part of the “Fresh Frozen Mouse Brain for Xenium Explorer Demo,” available at [10x Genomics](https://www.10xgenomics.com/datasets/fresh-frozen-mouse-brain-for-xenium-explorer-demo-1-standard). 

```{r data_load, message=FALSE, warning=FALSE, results='hide'}
## Load example ST data
# It contains a Seurat object
tf <- tempfile(fileext = ".rds")
utils::download.file("https://ndownloader.figshare.com/files/66021356", tf, mode = "wb")
xenium.obj <- readRDS(tf)
unlink(tf)

xenium.obj <- UpdateSeuratObject(xenium.obj)
xenium.obj$ccf_region <- gsub("1|2/3|4|5|6a|6b", "", xenium.obj$ccf) %>%
  as.factor() %>%
  droplevels()

# Subset celltype L2/3 IT
celltype <- "L2/3 IT"
xenium_sub.obj <- xenium.obj[, xenium.obj$ensembled.celltype1 %in% celltype]
```

Now we prepare the input data for **spCorr** analysis.

```{r data_prepare1}
# Count matrix
count_mat <- xenium_sub.obj[["Xenium"]]$counts
# Covariate matrix
cov_mat <- cbind(GetTissueCoordinates(xenium_sub.obj), xenium_sub.obj@meta.data)
rownames(cov_mat) <- colnames(count_mat)
```

Next we will use **MorphoGAM** ([Nicol *et al.*, 2024](https://doi.org/10.1101/2024.11.21.624653)) to find the 1D spatial curve associated with the 2D spatial coordinates of the spots. The 1D curve will be used as a covariate in the **spCorr** analysis.

```{r data_prepare2, fig.width=8, fig.height=4, fig.align='center'}
# Use MorphoGAM to find the 1D spatial curve
MorphoGAM_fit <- MorphoGAM::CurveFinder(cov_mat[, c("x", "y")] |> as.matrix(), knn = 5)
cov_mat$t <- MorphoGAM_fit$xyt$t

# (and plot)
df <- MorphoGAM_fit$xyt %>% arrange(t)
plot_curve <- ggplot(df, aes(x = x, y = y)) +
  geom_point(color = "gray70") +
  geom_path(aes(x = f1, y = f2, color = t)) +
  labs(title = paste(celltype, ": 1D spatial curve from MorphoGAM")) +
  scale_color_viridis_c(option = "magma") +
  coord_fixed() +
  scale_y_reverse() +
  theme_minimal()
plot_curve
```

In this tutorial, we will focus on a subset of 10 genes that are highly expressed in the L2/3 IT cell type. We will analyze all possible gene pairs among these 10 genes.

```{r data_prepare3}
gene_list <- c("Cux2", "Kcnh5", "Car4", "Rims3", "Bhlhe22", "Neurod6", "Gsg1l", "Cpne8", "Dpy19l1", "Cdh6")
gene_pair_list <- t(combn(gene_list, 2))
rownames(gene_pair_list) <- apply(gene_pair_list, 1, paste0, collapse = "_")
```

## Run spCorr

We now apply **spCorr** to infer spot-level gene–gene correlations for the selected gene pairs and to identify gene pairs with spatially varying correlation (SVC) pattern across the 1D spatial curve.

```{r run_spCorr}
res <- spCorr(count_mat,
  gene_list,
  gene_pair_list,
  cov_mat,
  formula1 = "1",
  formula2 = "s(t, bs='tp', k=8)",
  return_pi = TRUE,
  ncores = 5,
  seed = 123
)
```



## Visualize spot-level correlations 

We next visualize the inferred spot-level correlations and SVC patterns for specific gene pairs. 

```{r p_corr_t, fig.width=8, fig.height=4, fig.align='center'}
# Use the top gene pairs with the smallest FDR values for visualization
top_pairs <- names(sort(res$fdr))[1:6]

# Assemble spot-level correlation estimates and their confidence intervals
rho_df_long <- purrr::map_dfr(top_pairs, function(gp) {
  tibble::tibble(
    t = cov_mat$t, gene_pair = gp,
    rho = as.numeric(res$res_local[gp, ]),
    rho_lower = as.numeric(res$res_local_pi[[gp]]$lower),
    rho_upper = as.numeric(res$res_local_pi[[gp]]$upper)
  )
}) %>%
  dplyr::mutate(
    rho_lower = pmax(rho_lower, -1),
    rho_upper = pmin(rho_upper, 1),
    gene_pair = factor(gene_pair, levels = top_pairs)
  ) %>%
  dplyr::arrange(gene_pair, t)

# Visualize the estimated spot-level correlations along the 1D spatial curve
p_corr_t <- ggplot(rho_df_long, aes(x = t, y = rho)) +
  geom_ribbon(aes(ymin = rho_lower, ymax = rho_upper, group = gene_pair), fill = "purple", alpha = 0.25, color = NA) +
  geom_line(aes(group = gene_pair), color = "purple", linewidth = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~gene_pair, nrow = 2) +
  scale_x_continuous(breaks = c(0, 0.5, 1), labels = c("0", "0.5", "1")) +
  scale_y_continuous(breaks = c(-1, 0, 1), limits = c(-1, 1)) +
  labs(x = "t", y = "Correlation") +
  theme_minimal() +
  theme(
    aspect.ratio = 1 / 2,
    legend.position = "none",
    panel.grid = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.2),
    strip.background = element_rect(color = "black", fill = NA, linewidth = 0.2),
    strip.text.x = element_text(face = "italic")
  )
p_corr_t
```



## Session information

```{r}
sessionInfo()
```
